Model Finding for Recursive Functions in SMT

### 2016-04-11 ijcar2016/body.tex
Fix in Section 3 about the E(\SIgma) theory T

\newcommand{\Mo}{{\mathcal{I}}}

@@

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I,$
each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_{\smash{1}}^\I \times \cdots \times\allowbreak \tau_n^\I \rightarrow\allowbreak \tau^\I$,
and each variable $x:\tau$ of $\svars{\Sigma}$ to an element of $\tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$, closed under variable reassignment
(i.e., for every $I \in \Mo$, every $\Sigma$-interpretation that differs
from $I$ only on the variables of $\svars \Sigma$ is also in $\Mo$).
A $\Sigma$-formula $\varphi$ is \emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{rep};
otherwise, it is \emph{$T$-unsatisfiable}\end{rep}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent}
if each $T$-entails the other.
If $T_1 = (\Sigma_1, \Mo_1)$ is a theory and $\Sigma_2$ is a signature
with $\sfuns{\Sigma_1} \cap \sfuns{\Sigma_2} = \emptyset$,
the \emph{extension of $T_1$ to $\Sigma_2$} is the theory $T = (\Sigma, \Mo)$ where
$\sfuns{\Sigma} = \sfuns{\Sigma_1} \cup \sfuns{\Sigma_2}$,
$\stypes{\Sigma} = \stypes{\Sigma_1} \cup \stypes{\Sigma_2}$,
and $\Mo$ is the set of all $\Sigma$-interpretations $\I$
whose $\Sigma_1$-reduct is a model of $T_1$.
We refer to the symbols of $\Sigma_2$ that are not in $\Sigma_1$ as \emph{uninterpreted}.
For the rest of the paper, \relax{we fix a theory $T = (\Sigma, \Mo)$
with uninterpreted symbols} constructed as above.

@@

For the rest of the section, let $\varphi$ be a $\Sigma$-formula
in definitional form with respect to $\sfundefs{\Sigma}$
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\phi'$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
We present a method that constructs an extended signature
$\extendsig{ \Sigma }$ and an $\extendsig{ \Sigma }$-formula $\varphi'$ such that
$\varphi'$ is $T$-satisfiable if and only if the $\Sigma$-formula $\varphi$ is
$T$-satisfiable---i.e., $\varphi$ and $\varphi'$ are \emph{equisatisfiable} (\emph{in $T$}).
The idea behind this translation is to use an uninterpreted type $\alpha_\con{f}$
to abstract the set of
\emph{relevant} input tuples for each defined function $\con{f}$ and restrict the
quantification of $\con{f}$'s definition to a single variable of this type.
Informally, the relevant input tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.
More precisely,
%suppose our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
for each $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau \in \sfundefs{\Sigma}$,
the extended signature $\extendsig{\Sigma}$ contains
%\begin{itemize}
%\item
an uninterpreted \emph{abstract type} $\fargtype{\con{f}}$ 
abstracting the Cartesian product $\tau_1 \times \cdots \times \tau_n$
and
%\item
$n$ uninterpreted \emph{concretization functions} $\fargx{\con{f}}{1} : \fargtype{\con{f}} \rightarrow \tau_1$, $\dotsc,$ $\fargx{\con{f}}{n} : \fargtype{\con{f}} \rightarrow \tau_n$.
%\end{itemize}
%The interpretation of uninterpreted type $\fargtype{\con{f}}$ will denote the elements (tuples) on which the function $\con{f}$ is applied.
%The role of the uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ will be discussed more in the following.

>>>>

%\newcommand{\Mo}{{\mathcal{I}}}
\newcommand{\Mo}{{\mathbf{I}}}

@@

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I,$
each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_{\smash{1}}^\I \times \cdots \times\allowbreak \tau_n^\I \rightarrow\allowbreak \tau^\I$,
and each variable $x:\tau$ of $\svars{\Sigma}$ to an element of $\tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$, closed under variable reassignment
(i.e., for every $\I \in \Mo$, every $\Sigma$-interpretation that differs
from $\I$ only on the variables of $\svars \Sigma$ is also in $\Mo$).
A $\Sigma$-formula $\varphi$ is \emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{rep};
otherwise, it is \emph{$T$-unsatisfiable}\end{rep}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent}
if each $T$-entails the other.
If $T_0 = (\Sigma_0, \Mo_0)$ is a theory and $\Sigma$ is a signature
with $\sfuns{\Sigma_0} \subseteq \sfuns{\Sigma}$ 
and $\stypes{\Sigma_0} \subseteq \stypes{\Sigma}$,
the \emph{extension of $T_0$ to $\Sigma$} is the theory $T = (\Sigma, \Mo)$ 
where $\Mo$ is the set of all $\Sigma$-interpretations $\I$
whose $\Sigma_0$-reduct is a model of $T_0$.
Note that $T$ is a \emph{conservative} extension of $T_0$, in the sense that
a $\Sigma_0$-formula is $T_0$-satisfiable if and only it is $T$-satisfiable.
We refer to the symbols of $\Sigma$ that are not in $\Sigma_0$ as \emph{uninterpreted}.
For the rest of the paper, \relax{we fix a theory $T = (\Sigma, \Mo)$
with uninterpreted symbols} constructed as above.

@@

For the rest of the section, let $\varphi$ be a $\Sigma$-formula
in definitional form with respect to $\sfundefs{\Sigma}$
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\phi'$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
We present a method that constructs an extended signature
$\extendsig{ \Sigma }$ and an $\extendsig{ \Sigma }$-formula $\varphi'$ 
that is $T'$-satisfiable if and only if $\varphi$ is $T$-satisfiable
where $T'$ is the extension of $T$ to $\extendsig{ \Sigma }$---i.e., $\varphi$ and $\varphi'$ 
are \emph{equisatisfiable} (\emph{in $T'$}).
Since $T'$ is a conservative extension of $T$, for notational simplicity
we will refer to it also as $T$ from now on.
The idea behind this translation is to use an uninterpreted type $\alpha_\con{f}$
to abstract the set of
\emph{relevant} input tuples for each defined function $\con{f}$ and restrict the
quantification of $\con{f}$'s definition to a single variable of this type.
Informally, the relevant input tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.
More precisely,
%suppose our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
for each $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau \in \sfundefs{\Sigma}$,
the extended signature $\extendsig{\Sigma}$ contains
%\begin{itemize}
%\item
an uninterpreted \emph{abstract type} $\fargtype{\con{f}}$ 
abstracting the Cartesian product $\tau_1 \times \cdots \times \tau_n$
and
%\item
$n$ uninterpreted \emph{concretization functions} $\fargx{\con{f}}{1} : \fargtype{\con{f}} \rightarrow \tau_1$, $\dotsc,$ $\fargx{\con{f}}{n} : \fargtype{\con{f}} \rightarrow \tau_n$.
%\end{itemize}
%The interpretation of uninterpreted type $\fargtype{\con{f}}$ will denote the elements (tuples) on which the function $\con{f}$ is applied.
%The role of the uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ will be discussed more in the following.
